---
title: "scores data analysis"
author: "jcaron01"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 7
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

# 1. Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyr)
library(textshape)
library(patchwork)
library(emmeans)
library(UpSetR)
library(ggbeeswarm)
library(karyoploteR)
library(GenomicRanges)
library(GenomeInfoDb)
library(pheatmap)
library(igraph)
library(ggraph)
library(cowplot)
library(gtools)
library(wesanderson)
library(ggrepel)
library(VennDiagram)
library(grid)
library(scales)

theme_set(
  theme_classic() +
    theme(
      legend.position = "top",
      plot.title = element_text(face = "bold", size = 14),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
)
```

# 2. Data import

```{r column-names}
# Column names for the raw BED-like CRE annotation files loaded below
col_names = c("chr","start","end","gene","present_in","n_clusters","distanceToTSS","annotation","GC.percent","width","scores")
```

## 2.2 Helper function to read score files

```{r read-helpers}
# Read a BED-like CRE annotation file and tag its conservation score column
# with the name of the underlying metric (phastCons or phylop)
read_cre = function(path, phast) {
  df = read.delim(path, header = FALSE)
  colnames(df) = col_names
  df$chr = as.factor(df$chr)
  df$annotation = as.factor(df$annotation)
  df = df %>% mutate(width = end-start)
  df[[paste0("scores_", phast)]] = as.numeric(df$scores)
  df$scores = NULL
  df$gene = as.factor(df$gene)
  df
}
```

## 2.3 Load the CRE table

```{r load-main}
# Load each score type separately, then merge them into a single table
phastcons = read_cre("~/Documents/scores/mus_phastcons_v2.bed", "phastcons")
phylop = read_cre("~/Documents/scores/mus_phyloP_v2.bed", "phylop")

mus_scores = phastcons %>%
  left_join(phylop)

rm(phastcons)
rm(phylop)
```

## 2.4 Load coverage information

```{r load-coverage}
cov = read.delim("~/Documents/scores/coverage_CREs.bed", header = FALSE) 
col_cov = c("chr", "start", "end", "gene", "clusters", "n_clusters",   "distanceToTSS", "annotation", "GC.percent",   "width","scores_phastcons", "n_features", "n_bases", "length", "coverage") 
colnames(cov) = col_cov
# Keep only the columns needed to join coverage onto the main score table
cov = cov[, c("chr","start","end","gene","coverage")]
cov$chr = as.factor(cov$chr)
cov$gene = as.factor(cov$gene)

mus = mus_scores %>%
  left_join(cov)

rm(mus_scores)
rm(cov)

# Drop annotation categories with not enough data
mus = mus %>% 
  filter(annotation != "Downstream") %>% 
  filter(annotation != "5_UTR") %>% 
  droplevels()
```

# 3. Creating new features

## 3.1 Annotation helper functions

```{r annotation-helpers}
# Number of comma-separated elements in a string
number_elements = function(x) if (!is.na(x)) length(trimws(strsplit(x, ",")[[1]])) else 0

# Detect clusters (md/mx collapsed) that appear more than once, i.e. pleiotropic
# across replicates of the same cluster
detect_pleiotropy = function(x) {
  elements  = trimws(strsplit(x, ",")[[1]])
  prefixes  = sub("\\-(md|mx)$", "", elements)
  pleio  = prefixes[duplicated(prefixes) | duplicated(prefixes, fromLast = TRUE)]
  if (length(pleio) == 0) return(NA)
  paste(unique(pleio), collapse = ", ")
}

# Extract only the elements that match a given md/mx suffix
extract_suffix = function(x, suffix) {
  elements = trimws(strsplit(x, ",")[[1]])
  sfx = sub("^(epi|mes|Endothelium)\\-[1-7]\\-", "", elements)
  sfx = sub("^(Endothelium)\\-", "", sfx)
  matched  = elements[sfx == suffix]
  if (length(matched) == 0) return(NA)
  paste(matched, collapse = ", ")
}

# Count how many elements of a comma-separated column match a regex pattern
count_type = function(col, pattern) {
  sapply(strsplit(col, ",\\s*"), function(x) sum(grepl(pattern, trimws(x))))
}

# Collapse replicate-level labels down to unique cluster names
detect_cluster = function(x) {
  paste(unique(sub("\\-(md|mx)$", "", trimws(strsplit(x, ",")[[1]]))), collapse = ", ")
}

# Remove specific clusters (e.g. noisy/low-quality ones) from a comma-separated list
filter_clusters = function(x, exclude = c("mes-3", "Endothelium")) {
  elements = trimws(strsplit(x, ",")[[1]])
  kept = elements[!(elements %in% exclude)]
  if (length(kept) == 0) return(NA)
  paste(kept, collapse = ", ")
}
```

## 3.2 Add new variables in the dataframe

```{r feature-engineering}
# Cluster membership, cleaned of excluded clusters
mus$clusters = sapply(mus$present_in, detect_cluster)
mus$clusters = sapply(mus$clusters, filter_clusters)
mus = mus %>%
  filter(!is.na(clusters))
mus$n_clusters = sapply(mus$clusters, number_elements)

# Pleiotropy: clusters found in more than one replicate
mus$pleio_clusters = sapply(mus$present_in, detect_pleiotropy)
mus$n_pleio = sapply(mus$pleio_clusters, number_elements)

# md / mx (lower / upper) subsets and their counts
mus$mx_clusters = sapply(mus$present_in, extract_suffix, suffix = "mx")
mus$md_clusters = sapply(mus$present_in, extract_suffix, suffix = "md")
mus$n_mx  = sapply(mus$mx_clusters, number_elements)
mus$n_md  = sapply(mus$md_clusters, number_elements)

# Epithelium / mesenchyme cluster counts
mus$n_epi = count_type(mus$clusters, "^epi\\-")
mus$n_mes = count_type(mus$clusters, "^mes\\-")

mus = mus %>% mutate(
  only_mx = n_mx > 0 & n_md == 0,
  only_md = n_md > 0 & n_mx == 0,
  only_pleio = ifelse(n_pleio > 0, n_pleio * 2 == n_mx + n_md, FALSE),
  only_epi = n_epi > 0 & n_mes == 0,
  only_mes = n_epi == 0 & n_mes > 0,
  spécificité = case_when(only_md | only_mx ~ "Modulaire",TRUE ~ "Pléiotrope"),
  conservation = ifelse(coverage > .5, "Conservé","Non conservé"),
  range_clusters = factor(
    case_when(n_clusters == 1 ~ "Specific", n_clusters == 2 ~ "2",
              n_clusters == 3 ~ "3", TRUE ~ ">3"),
    levels = c("Specific", "2", "3", ">3")
  )
)

# Natural sort of chromosome names instead of lexical order
mus$chr = factor(mus$chr, levels = mixedsort(levels(mus$chr)))
```

# 4. Exploratory analysis

## 4.1 Missing data and score summaries

```{r na-summary}
# Report NA counts and summary statistics for each conservation score column
for (score_var in c("scores_phastcons", "scores_phylop")) {
  print(paste(score_var,"info:"))
  print("#NA:")
  print(table(is.na(mus[[score_var]])))
  print("Summary:")
  print(summary(mus[[score_var]]))
}
```

## 4.2 Peak classification counts

```{r classification-counts}
# CRE count per annotation category, sorted from most to least common
mus %>%
  count(annotation, name = "n") %>%
  arrange(desc(n))
```

# 5. Linear models

```{r model-fitting-function}
# Fit a linear model with stepwise term selection, then print its summary and ANOVA table
fit_model = function(formula_str, data) {
  m = step(lm(as.formula(formula_str), data = data))
  print(summary(m))
  print(anova(m))
  m
}
```

## 5.1 phastCons model

```{r model-phastcons}
model_phastcons = fit_model("scores_phastcons ~ n_clusters*annotation*width", mus)
plot(emmeans(model_phastcons, ~ annotation), comparisons = TRUE)
```

## 5.2 phylop model

```{r model-phylop}
model_phylop = fit_model("scores_phylop ~ n_clusters*annotation*width", mus)
plot(emmeans(model_phylop, ~ annotation), comparisons = TRUE)
```

## 5.3 Other models

```{r model-clusters-vs-annotation}
# Does the number of clusters differ significantly across annotation categories?
anova(lm(n_clusters ~ annotation, mus))

#Are PhastCons and PhyloP scores correlated ? 
cor(mus$scores_phastcons, mus$scores_phylop, use = "pairwise.complete.obs")
```

# 6. Summary statistics by classification

```{r summary-by-classification}
# Mean, median, sd of each score, plus sample size, per annotation category
mus %>%
  group_by(annotation) %>%
  summarise(across(c(scores_phastcons, scores_phylop),
                    list(mean = ~ mean(.x, na.rm = TRUE),
                         median = ~ median(.x, na.rm = TRUE),
                         sd = ~ sd(.x, na.rm = TRUE)),
                    .names = "{.fn}_{.col}"),
            n = n())
```

# 7. Plotting functions

```{r plotting-functions}
# Pie chart of the distribution of a categorical variable
classif_pie = function(df, fill_var, title,
                        palette = scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous"))) {

  dat = as.data.frame(table(df[[fill_var]]))
  colnames(dat) = c("groupe", "n")
  dat$pct   = round(dat$n / sum(dat$n) * 100, 1)
  dat$label = paste0(dat$groupe, "\n", dat$pct, "%")

  ggplot(dat, aes(x = "", y = n, fill = groupe)) +
    geom_col(width = 1, color = "white") +
    palette +
    coord_polar(theta = "y") +
    # tous les labels SAUF 3_UTR, positionnement normal
    geom_label(data = subset(dat, groupe != "3_UTR"),
               aes(label = label), position = position_stack(vjust = 0.5),
               size = 3, show.legend = FALSE) +
    # uniquement 3_UTR, décalé pour éviter le chevauchement
    geom_label(data = subset(dat, groupe == "3_UTR"),
               aes(x = 1, y = .8, label = label),
               size = 2.8, show.legend = FALSE, inherit.aes = FALSE,
               fill = "#3B8EA5", color = "black") +
    labs(title = title, fill = fill_var) +
    theme_void() +
    theme(legend.position = "none", title = element_text(face = "bold"))
}

# Mean score (with bootstrap CI) by a grouping variable, annotated with sample size
score_point = function(df, x, score_var, title, x_label, y_label, facet = FALSE, fill = "annotation", palette = scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous"))) {
  n_fun = function(y) {
    data.frame(y = mean(y, na.rm = TRUE)*1.1, label = paste0("n=", length(y)))
  }
  
  p = ggplot(df, aes(x = .data[[x]], y = .data[[score_var]], fill = .data[[fill]])) +
    stat_summary(fun.data = mean_cl_boot, geom = "pointrange",
                 shape = 21, color = "black", size = 0.8) +
    stat_summary(fun.data = n_fun, geom = "text", size = 3, color = "black") +
    palette +
    labs(title = title, x = x_label, y = y_label) + 
    theme(legend.position = "None")
  if (facet) p = p + facet_wrap(~ annotation, scales = "free")
  p
}

# Mean score (with bootstrap CI) as a function of the number of clusters
score_by_cluster_pointrange = function(df, x_var, score_var, color, title, y_lab, facet = FALSE) {
  p = ggplot(df, aes(x = factor(.data[[x_var]]), y = .data[[score_var]])) +
    stat_summary(fun.data = mean_cl_boot, geom = "pointrange", size = 0.8, color = color,
                 fun.args = list(conf.int = 0.95, B = 1000)) +
    labs(title = title, x = "Nombre de types cellulaires", y = y_lab)
  if (facet) p = p + facet_wrap(~ annotation, scales = "free")
  p
}

# Generic boxplot of a score variable by any categorical variable, faceted by classification
score_boxplot_facet = function(df, x_var, score_var, title) {
  n_fun = function(x) {
    data.frame(y = max(x) * 1.05, label = paste0("n=", length(x)))
  }
  ggplot(df, aes(x = .data[[x_var]], y = .data[[score_var]], fill = .data[[x_var]])) +
    geom_boxplot(alpha = 0.7) +
    stat_summary(fun.data = n_fun, geom = "text", size = 3) +
    facet_wrap(~ annotation, scales = "free") +
    scale_fill_manual(values = wes_palette("Cavalcanti1")) +
    labs(title = title) +
    theme(legend.position = "none")
}
```

# 8. Descriptive plots

## 8.1 Overall distribution of CRE classifications

```{r plot-classification-pie}
pie_annotation = classif_pie(mus, "annotation", "Distribution des région régulatrices")
pie_annotation
```

## 8.2 Score distributions, conservation, and pleiotropy per classification

```{r plot-score-distributions, fig.height=8}
pleio_vs_specif = score_point(mus, "spécificité", "scores_phastcons", "Scores PhastCons moyens:\n Modulaire vs Pléiotrope", "Spécificité", "PhastCons moyen", facet = T, fill = "spécificité", palette = scale_fill_manual(values = wes_palette("GrandBudapest2")))
scores_vs_anno = score_point(mus, "annotation", "scores_phastcons", "Scores PhastCons moyens par annotation", "Annotation", "PhastCons moyen")
scores_vs_cons = score_boxplot_facet(mus, "conservation", "scores_phastcons", paste("Score Phastcons: conservé vs non-conservé par annotation"))
pleio_vs_specif
scores_vs_cons
scores_vs_anno
```

## 8.3 Clusters per classification

```{r plot-clusters-per-classification}
ggplot(mus, aes(x = factor(n_clusters), fill = annotation)) +
  geom_bar(position = "dodge") + scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous")) + labs(title = "Clusters per classification")
```

## 8.4 CREs per chromosome

```{r plot-cre-per-chromosome}
cre_per_chrom = ggplot(mus, aes(x = chr, fill = annotation)) +
  geom_bar(position = "fill") + scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous")) + labs(title = "Annotations des régions régulatrices par chromosome", x = "Chromosome", y = "Pourcentage")
cre_per_chrom
```

## 8.5 Conservation per classification

```{r plot-conservation-per-classification}
cons_vs_anno = ggplot(mus, aes(x = annotation, fill = conservation)) +
    geom_bar(position = "fill") +
    scale_fill_manual(values = wes_palette("Cavalcanti1")) +
    labs(title = "Conservation par classification", y = "Pourcentage")
cons_vs_anno
```

## 8.7 Mean conservation score by number of clusters

```{r plot-score-by-cluster}
scores_vs_n_clusters = score_by_cluster_pointrange(mus, "n_clusters", "scores_phastcons", "red4", "Score PhastCons moyen par nombre de types cellulaires", "PhastCons moyen")
scores_vs_n_clusters
```

## 8.8 Mean number of clusters per classification

```{r plot-mean-clusters}
mean_clusters = ggplot(mus, aes(x = annotation, y = n_clusters, fill = annotation)) +
  stat_summary(fun = mean, geom = "bar") +
  scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous")) +
  labs(title = "Nombre moyen de types cellulaires par annotation", x = "Annotation", y = "Moyenne") +
  theme(legend.position = "None")
mean_clusters
```

## 8.9 Conservation score vs. CRE width (binned)

```{r plot-score-vs-width}
# Bin CREs into 10 equal-sized width quantiles, then average each score per bin
mus_num = mus %>%
  select(width, scores_phastcons, scores_phylop) %>%
  mutate(bin = cut_number(width, n = 10))

bins = mus_num %>%
  group_by(bin) %>%
  summarise(
    width = mean(width, na.rm = TRUE),
    scores_phastcons = mean(scores_phastcons, na.rm = TRUE),
    scores_phylop    = mean(scores_phylop, na.rm = TRUE),
    .groups = "drop"
  )

colors = c(scores_phastcons = "red4", scores_phylop = "purple4")
nice_labels = c(scores_phastcons = "Score PhastCons", scores_phylop = "PhyloP score")

plots = list()
for (sv in c("scores_phastcons", "scores_phylop")) {
  p = ggplot(bins, aes(x = bin, y = .data[[sv]], group = 1)) +
    geom_line(color = colors[[sv]], linewidth = 1, alpha = 0.5) +
    geom_point(shape = 21, size = 3.5, fill = colors[[sv]], color = "white",
               stroke = 0.6) +
    labs(
      title  = paste(nice_labels[[sv]], "en fonction de la taille de la séquence"),
      x = "Bins de taille (bp)",
      y = nice_labels[[sv]]
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
  plots[[sv]] = p
  print(p)
}

# Same binning, but split out by annotation category
mus_num = mus %>%
  select(width, scores_phastcons, scores_phylop, annotation) %>%
  mutate(bin = cut_number(width, n = 10))

bins = mus_num %>%
  group_by(annotation, bin) %>%
  summarise(
    width = mean(width, na.rm = TRUE),
    scores_phastcons = mean(scores_phastcons, na.rm = TRUE),
    scores_phylop    = mean(scores_phylop, na.rm = TRUE),
    .groups = "drop"
  )

colors = c(scores_phastcons = "red4", scores_phylop = "purple4")
nice_labels = c(scores_phastcons = "PhastCons score", scores_phylop = "PhyloP score")

plots_facet = list()
for (sv in c("scores_phastcons", "scores_phylop")) {
  p = ggplot(bins, aes(x = bin, y = .data[[sv]], group = 1)) +
    geom_line(color = colors[[sv]], linewidth = 1, alpha = 0.5) +
    geom_point(shape = 21, size = 3.5, fill = colors[[sv]], color = "white",
               stroke = 0.6) +
    facet_wrap(~annotation, scales = "free_y") +
    labs(
      title = paste(nice_labels[[sv]], "vs width"),
      subtitle = "Mean by bin (15 quantiles)",
      x = "Width bins (bp)",
      y = nice_labels[[sv]]
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40", size = 10, margin = margin(b = 10)),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
  plots_facet[[sv]] = p
  print(p)
}
scores_vs_width = plots$scores_phastcons
scores_vs_width_facet = plots_facet$scores_phastcons
```

## 8.10 Genomic density of CREs by chromosome

```{r plot-density-by-chr}
density = ggplot(mus, aes(x = start, fill = chr)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ chr, scales = "free") +
  scale_fill_manual(values = wes_palette("Zissou1", n = 21, type = "continuous")) +
  labs(title = "Density of CREs by Chromosome") +
  theme(legend.position = "none")
density
```

# 9. Pleiotropy and tissue-specificity pie charts

## 9.1 Pleiotropy (mx vs md vs pleiotropic)

```{r plot-pleio-pie}
pie_pleio = classif_pie(mus, "spécificité","Distribution de la pléiotropie", palette = scale_fill_manual(values = wes_palette("GrandBudapest2")))

pie_pleio
```

## 9.2 Tissue specificity (epithelium vs mesenchyme)

```{r plot-tissue-pie}
mus |>
  count(only_epi, only_mes) |>
  mutate(
    label = case_when(only_epi ~ "only epi", only_mes ~ "only mes", TRUE ~ "not specific"),
    pct   = round(n / sum(n) * 100, 1)
  ) |>
  ggplot(aes(x = "", y = n, fill = label)) +
  geom_col(width = 1, color = "white", linewidth = 0.5) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(pct, "%")), position = position_stack(vjust = 0.5), size = 4, fontface = "bold") +
  scale_fill_hue() +
  labs(title = "Data distribution of tissues", fill = NULL) +
  theme_void() +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

## 9.3 Pleiotropy (specific vs not) per classification

```{r plot-pleio-per-classification}
pleio_per_anno = ggplot(mus, aes(x = annotation, fill = spécificité)) +
  geom_bar(position = "fill") + scale_fill_manual(values = wes_palette("GrandBudapest2")) + labs(title = "Pléiotropie par annotation", y = "Pourcentage")
pleio_per_anno
```

# 11. Replicate-level analysis within the epithelium

```{r epi-replicate-function}
# For a given epithelium replicate i, classify each CRE as pleiotropic
# (present in both md/mx forms) or modular (present in only one)
analyze_epi = function(df, i) {
  epi_col = count_type(df$present_in, paste0("^epi-", i, "\\-"))
  df %>%
    mutate(
      in_epi = epi_col > 0,
      pleio_epi = epi_col == 2,
      condition = case_when(pleio_epi ~ "pleio", in_epi & !pleio_epi ~ "modular")
    ) %>%
    filter(!is.na(condition))
}
```

```{r epi-replicate-combine}
# Run the epithelium classification across all 7 replicates and combine the results
epi_results_combined = bind_rows(
  lapply(1:7, function(i) {
    d = analyze_epi(mus, i)
    if (nrow(d) > 0) mutate(d, epi = paste0("epi-", i)) else NULL
  })
) %>%
  mutate(annotation = as.factor(annotation)) %>%
  filter(!annotation %in% "5_UTR") %>%
  mutate(annotation = droplevels(annotation))
```

```{r epi-replicate-plot-function}
# Bar plot (mean +/- bootstrap CI) of a score by condition, faceted by
# epithelium replicate, annotation, and cluster-count range
epi_bar_plot = function(df, score_var, y_label, title) {
  ggplot(df, aes(x = condition, y = .data[[score_var]], fill = annotation)) +
    stat_summary(fun = mean, geom = "bar") +
    stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.25, linewidth = 0.5) +
    scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous")) +
    facet_grid(epi ~ annotation + range_clusters, scales = "free") +
    labs(x = "Condition", y = y_label, title = title) +
    theme(legend.position = "none")
}
```

```{r epi-replicate-plots, fig.width=14, fig.height=12}
epi_bar_plot(epi_results_combined, "scores_phastcons", "Phastcons score",
             "Phastcons vs condition by epi and number of clusters")
epi_bar_plot(epi_results_combined, "scores_phylop",    "phylop score",
             "phylop vs condition by epi and number of clusters")
```

# 11.2 Replicate-level analysis within the mesenchyme

```{r mes-replicate-function}
# Same logic as analyze_epi(), applied to mesenchyme replicates
analyze_mes = function(df, i) {
  mes_col = count_type(df$present_in, paste0("^mes-", i, "\\-"))
  df %>%
    mutate(
      in_mes    = mes_col > 0,
      pleio_mes = mes_col == 2,
      condition = case_when(pleio_mes ~ "pleio", in_mes & !pleio_mes ~ "modular")
    ) %>%
    filter(!is.na(condition))
}
```

```{r mes-replicate-combine}
# Run the mesenchyme classification across all 7 replicates, dropping the
# noisy mes-3 replicate, and combine the results
mes_results_combined = bind_rows(
  lapply(1:7, function(i) {
    d = analyze_mes(mus, i)
    if (nrow(d) > 0) mutate(d, mes = paste0("mes-", i)) else NULL
  })
) %>%
  mutate(annotation = as.factor(annotation)) %>%
  filter(!annotation %in% c("Downstream", "5_UTR")) %>%
  mutate(annotation = droplevels(annotation)) %>%
  filter(mes != "mes-3")
```

```{r mes-replicate-plot-function}
# Same bar-plot logic as epi_bar_plot(), faceted by mesenchyme replicate instead
mes_bar_plot = function(df, score_var, y_label, title) {
  ggplot(df, aes(x = condition, y = .data[[score_var]], fill = annotation)) +
    stat_summary(fun = mean, geom = "bar") +
    stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.25, linewidth = 0.5) +
    scale_fill_manual(values = wes_palette("Zissou1", n = 5, type = "continuous")) +
    facet_grid(mes ~ annotation + range_clusters, scales = "free") +
    labs(x = "Condition", y = y_label, title = title) +
    theme(legend.position = "none")
}
```

```{r mes-replicate-plots, fig.width=14, fig.height=12}
mes_bar_plot(mes_results_combined, "scores_phastcons", "Phastcons score",
             "Phastcons vs condition by mes and number of clusters")
mes_bar_plot(mes_results_combined, "scores_phylop",    "phylop score",
             "phylop vs condition by mes and number of clusters")
```

# 12. UpSet plots

```{r upset-function}
# Build an UpSet plot from a list of logical membership vectors
make_upset = function(df, sets_list, sets_names) {
  d = as.data.frame(setNames(lapply(sets_list, as.integer), sets_names))
  UpSetR::upset(d, sets = sets_names, order.by = "freq", text.scale = 1.3,
        mainbar.y.label = "Number of CREs", sets.x.label = "Total",
        main.bar.color = "#443983", sets.bar.color = "#31688e", matrix.color = "#35b779")
}
```

## 12.1 Tissue overlap

```{r upset-tissue}
make_upset(mus,
           list(mus$n_epi > 0, mus$n_mes > 0),
           c("Epithelium", "Mesenchyme"))
```

## 12.2 md / mx / pleiotropy overlap

```{r upset-mdmx}
make_upset(mus,
           list(mus$n_mx > 0, mus$n_md > 0, mus$n_pleio > 0),
           c("mx", "md", "pleio"))
```

## 12.3 Conservation / specificity / pleiotropy overlap

```{r upset-cons}
make_upset(mus,
           list(mus$conservation == "Conservé", mus$only_md, mus$only_mx, mus$n_pleio > 0),
           c("conserved", "only_md", "only_mx", "pleio"))
```

# 13. Plots

## 13.1 Figures for internship report

```{r plot-rapport}
# Combine individual plots from previous sections into multi-panel report figures
fig_ab = cowplot::plot_grid(pie_annotation, mean_clusters, ncol = 2, labels = c("a", "b"), label_size = 30)
fig_1 = cowplot::plot_grid(fig_ab, cre_per_chrom, nrow = 2, labels = c("","c"), label_size = 30)
fig_1

fig_2 = cowplot::plot_grid(pie_pleio, pleio_per_anno, ncol = 2, labels = c("a", "b"), label_size = 30)
fig_2

fig_3 = cowplot::plot_grid(scores_vs_anno, pleio_vs_specif, scores_vs_n_clusters,scores_vs_width, labels = c("a","b","c","d"), label_size = 30)
fig_3

fig_4 = cowplot::plot_grid(cons_vs_anno, scores_vs_cons, nrow = 2, labels = c("a","b"), label_size = 30)
fig_4
```

## 13.2 Venn diagram

```{r plot-venn-mouse-hamster}
souris_total = 265138
hamster_total = 167268
intersection = 66353

#Exclusive zones
souris_seule = souris_total - intersection
hamster_seul = hamster_total - intersection

# Création du diagramme
draw.pairwise.venn(
  area1 = souris_total,
  area2 = hamster_total,
  cross.area = intersection,
  fill = c("#4C72B0", "#DD8452"),       
  alpha = 0.7,                          
  lty = "blank",                        
  cex = 1.5,                            
  fontfamily = "sans",
  cat.cex = 1.6,                       
  cat.fontface = "bold",
  cat.fontfamily = "sans",
  cat.pos = c(-20, 20),                 
  cat.dist = c(0.05, 0.05),
  scaled = TRUE,                        
  euler.d = TRUE,
  title = "Intersection entre les séquences du hamster et de la souris"
)
```

## 13.3 liftOver information

```{r plot-liftover-before-after}
df = data.frame(
  espece      = c("Souris", "Hamster"),
  version_av  = c("mm39", "MesAur2.0"),
  version_ap  = c("mm38", "MesAur1.0"),
  avant       = c(268325, 264514),
  apres       = c(265138, 167268)
)

df$pct_perdu = 100 * (df$avant - df$apres) / df$avant

ggplot(df, aes(x = espece)) +
  geom_col(aes(y = avant, fill = version_av), width = 0.6) +
  geom_col(aes(y = apres, fill = version_ap), width = 0.35) +
  geom_text(aes(y = avant, label = format(avant, big.mark = " ")),
            vjust = -.5, size = 4) +
  geom_text(aes(y = apres, label = format(apres, big.mark = " ")),
            vjust = 2, size = 4) +
  geom_text(aes(y = avant, label = paste0("-", round(pct_perdu, 1), " %")),
            vjust = -2.2, size = 5, fontface = "bold", color = "darkred") +
  scale_fill_manual(
    values = c("mm39" = "#4C72B0", "mm38" = "#A9C4EB",
               "MesAur2.0" = "#DD8452", "MesAur1.0" = "#F3C6A6"),
    name = "Version du génome"
  ) +
  labs(
    title = "Nombre de données avant/après liftOver pour la souris et le hamster",
    subtitle = "Barres claires: valeurs après liftOver",
    x = NULL, y = "Nombre de séquences cis-régulatrices"
  ) +
  ylim(0, max(df$avant) * 1.3)
```

## 13.4 Orthologous regions pie chart

```{r plot-orthologous-region-pie}
data = data.frame(
  categorie = c("Distal_Intergenic", "Downstream", "Exon", "Intron", "Promoter"),
  effectif = c(37873, 77, 3180, 22307, 2916)
)

data$pourcentage = data$effectif / sum(data$effectif) * 100
data$label = paste0(data$categorie, "\n", scales::comma(data$effectif), 
                      " (", round(data$pourcentage, 1), "%)")

data$categorie = factor(data$categorie, levels = data$categorie[order(-data$effectif)])

palette_zissou = wes_palette("Zissou1", n = 5, type = "continuous")

p = ggplot(data, aes(x = "", y = effectif, fill = categorie)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = palette_zissou) +
  geom_text(aes(label = paste0(round(pourcentage, 1), "%")),
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold") +
  labs(
    title = "Distribution des régions orthologues",
    fill = "Catégorie"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  )

print(p)
```
